V4 Neural Network Model for Shape-based Feature Extraction and Object Discrimination

### 2015-09-07 article.tex
fix typo

Understanding the content of images has always been a difficult task in image analysis due to the well known semantic gap \cite{smeulders2000}.
A high degree of abstraction is required to transform low-level pixel data into high-level knowledge of the images.
In an attempt to bridge this semantic gap, various methods have been proposed to extract features from the pixel representation of images.
However, humans and primates outperform any existing computer vision system with respect to almost any measure.
Though the neural mechanism of vision has not been fully revealed,
recent research on neuroscience do provide many inspiring discoveries which would help development of computer vision algorithms.
In this paper, we demonstrate a novel model of visual area V4 and its application to shape-based feature extraction and object discrimination.
Experiment results show that the proposed model exhibits similar shape selectivity to that of real V4 neurons
and it outperforms many existing methods in computer vision tasks including feature matching and object recognition.

@@

In this paper, we propose a multilayer neural network model which imitates the neural mechanism of visual area V4 in the ventral visual pathway.
Different stages of the neural pathway in the visual cortex are explicitly mapped to different layers of the neural network model.
In this feed-forward network, features are extracted hierarchically from input images.
High-level local image features emerge in the output layer of the network. 
The features encode local structures of object shapes in a way which closely resembles the mechanism of the cortex.
With a limited number of training images, the model can learn to distinguish between different categories of objects.
Moreover, this multilayer neural network shows a successful attempt to integrate different stages of visual processing in the brain. 
It provides an extensible framework so that future findings in cortex mechanisms can be adopted and applied to computer vision tasks.

We briefly introduce some related work on image features and V4 models in section~\ref{sec:2}.
In section~\ref{sec:3}, we give the detailed description of our proposed model.
It is capable of locating salient points and generating discriminative local shape features around salient points.
In section~\ref{sec:4}, we show several experiments in which we evaluate the proposed model.
The model is compared with real V4 neurons in a shape distinction experiment.
The features extracted by the model are applied to various computer vision tasks.
The discussion and conclusion is summarized in section~\ref{sec:5}.

@@

Complex cells are modeled as Gaussian filters which take inputs from simple cells,
\begin{equation}
f(x,y;\sigma_c)=\frac{1}{2\pi\sigma_c^2}\exp\left(-\frac{x^2+y^2}{2\sigma_c^2}\right),
\end{equation}
where $\sigma_c$ represents the radius of the complex receptive fields.
The radius of complex receptive field is approximately twice as long as that of a simple cell.
We take this size because qualitatively the complex receptive fields are usually two to five times wider 
than the bars of gratings that stimulate them most effectively \cite{hubel1962}.
In this paper, we have $\sigma_c=2\sigma_s$.

>>>>

Understanding the content of images has always been a difficult task in image analysis due to the well known semantic gap \cite{smeulders2000}.
A high degree of abstraction is required to transform low-level pixel data into high-level knowledge of the images.
In an attempt to bridge this semantic gap, various methods have been proposed to extract features from the pixel representation of images.
However, humans and primates outperform any existing computer vision system with respect to almost any measure.
Though the neural mechanism of vision has not been fully revealed,
recent research on neuroscience do provide many inspiring discoveries which would help development of computer vision algorithms.
In this paper, we demonstrate a novel model of visual area V4 and its application to shape-based feature extraction and object discrimination.
Experimental results show that the proposed model exhibits similar shape selectivity to that of real V4 neurons
and it outperforms many existing methods in computer vision tasks including feature matching and object recognition.

@@

In this paper, we propose a multilayer neural network model which imitates the neural mechanism of visual area V4 in the ventral visual pathway.
Different stages of the neural pathway in the visual cortex are explicitly mapped to different layers of the neural network model.
In this feed-forward network, features are extracted hierarchically from input images.
High-level local image features emerge in the output layer of the network. 
The features encode local structures of object shapes in a way which closely resembles the mechanism of the cortex.
With a limited number of training images, the model can learn to distinguish between different categories of objects.
Moreover, this multilayer neural network shows a successful attempt to integrate different stages of visual processing in the brain. 
It provides an extensible framework so that future findings in cortical mechanisms can be adopted and applied to computer vision tasks.

We briefly introduce some related work on V4 models in section~\ref{sec:2}.
In section~\ref{sec:3}, we give the detailed description of the proposed model.
It is capable of locating salient points and generating discriminative local shape features around salient points.
In section~\ref{sec:4}, we show several experiments in which we evaluate the model.
The model is compared with real V4 neurons in a shape distinction experiment.
The features extracted by the model are applied to various computer vision tasks.
The discussion and conclusion is summarized in section~\ref{sec:5}.

@@

Complex cells are modeled as Gaussian filters which take input from simple cells,
\begin{equation}
f(x,y;\sigma_c)=\frac{1}{2\pi\sigma_c^2}\exp\left(-\frac{x^2+y^2}{2\sigma_c^2}\right),
\end{equation}
where $\sigma_c$ represents the radius of the complex receptive fields.
The radius of complex receptive field is approximately twice as long as that of a simple cell.
We take this size because qualitatively the complex receptive fields are usually two to five times wider 
than the bars of gratings that stimulate them most effectively \cite{hubel1962}.
In this paper, we have $\sigma_c=2\sigma_s$.
